Hydrodynamics and rheology of adhesive vesicle doublets

### 2019-02-13 vesicle_code/docs/adhesion/adhesion.tex
small changes and combined appendicies

Adhesion potential for a vesicle membrane has been modeled by a L.-J. type potential
%Sukumaran and Seifert~\cite{suk-sei2001} consider adhesion between a
%single three-dimensional vesicle and a solid wall at $z=0$.  They use
%the adhesion potential of the function form
\begin{align}
\label{eq:adhesion_potential}
  \phi(z) = \mathcal{H} \left[ 
    \left(\frac{\delta}{z}\right)^m - \frac{m}{n} \left(\frac{\delta}{z}\right)^n \right],
\end{align}
where $\mathcal{H}$ is the Hamaker constant and $\delta$ is the adhesion length
scale. $z$ is the distance between a patch on the vesicle membrane and a point on the other object,
which could be another vesicle membrane or a flat solid wall.  The characteristics of the L.-J. potential are summarized in
the left panel of Figure~\ref{fig:adhesionModel}: The adhesive force is zero at the equilibrium distance $z=\delta$.
For large distances ($z> \delta$) the adhesive potential is attractive while for small distances ($z < \delta$) the potential is repulsive to prevent physical contact.

The exponents $(m,n)$ depend on the geometry and the molecular details of the two objects under
adhesion~\cite{Book_IntermolecularSurfaceForces}: $(m,n)=(4,2)$
corresponds to two flat, planar surfaces interacting with each other, and has been a common choice 
for studying membrane-solid adhesion  \cite{Seifert1990_PRA,suk-sei2001,ShiFengGao2006_ActaMechSin, LinFreund2007_IntJSolidsStructures,BlountMiksisDavis2013_PRSa,YoungStone2017_PRF}.
%
On the other hand,
$(m,n) = (12,6)$ corresponds to the Lennard-Jones (L.-J.)~potential
between two molecules, and has been used to model membrane-membrane adhesion  \cite{FlormannAouane2017_SciReports}.  
In our case, the adhesion potential in the
integral of equation~(\eqref{eqn:adhesionForce}) is between two small
patches of lipid bilayer membrane (because the lipid molecules are
coarse-grained in the continuum modeling).  
Thus a reasonable choice for
$(m,n)$ between two coarse-grained membrane patches would be between
those for two planar surfaces and two point molecules.

Large value of $m$ corresponds to a sharp increase in the repulsion
force as two objects are within the separation distance $\delta$.  This
poses a numerical challenge since the problem becomes stiff (i.e.,
requires a very small time step) for large $m$.  The adaptive
time-stepping BIE scheme makes it possible to simulate vesicle adhesive
dynamics with specified numerical precision for reasonable computation
time.  We explore several combinations of $(m,n)$ in the simulations of
two vesicles forming a doublet in a quiescent flow and in a shear flow.  We found that, as long as the
close-range interaction is well resolved both in space (using the near-singular evaluations) 
and time (using the high-order adaptive time integration), there is very
little difference in both the dynamic evolution and equilibrium
configuration between $(m,n)=(8,6)$ and $(m,n)=(4,2)$.  Therefore, in
this work, we use $(m,n) = (4,2)$ to regulate the numerical stiffness introduced
by the adhesive force. 
%
%For some simulations we used both sets of $(m,n)$ and did not notice significant difference
%in terms of adhesive hydrodynamics of vesicles.

Focusing on intermediate adhesive strengths, 
we use Hamaker constants ranging from $\mathcal{O}(10^{-1})$ to
$\mathcal{O}(10^0)$ times the bending rigidity modulus.  
We assume that the adhesion force from equation~\ref{eq:adhesion_potential} applies
between all pairs of points on different vesicles, and in
Appendix~\ref{sec:appendixA} the net adhesive force at a point $\xx$ on
vesicle $j$ is shown to be
\begin{align}
  \AA\xx:=-\mathcal{H} m \delta^{n}\sum_{\substack{k=1 \\ k \neq j}}^M 
  \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^{m+2}} 
  \left(\delta^{m-n} - \|\xx - \yy\|^{m-n} \right) ds_\yy.
  \label{eqn:adhesionForce}
\end{align}
This summation of adhesive forces is illustrated in the middle panel of Figure~\ref{fig:adhesionModel}. 
The right panel of Figure~\ref{fig:adhesionModel} is an 
example of the calculated adhesive force projected onto their center-of-mass vector. We notice that 
the adhesive force is repulsive in the contact region, while for the rest of the membranes there is an attractive force
that acts to pull the two vesicles together. The summary of boundary integral formulation can be found in Appendix \ref{sec:AppendixB} and
Appendix \ref{sec:AppendixC}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Shear and Normal Stresses}
\label{sec:AppendixC}
To characterize the rheological properties of the suspension of a
doublet, it is necessary to compute the shear and normal stresses.  At a
point $\xx \in \RR^2 \backslash \gamma$ in the fluid bulk, the velocity
$\uu(\xx)$, pressure $p(\xx)$, and total stress tensor $T = - p I +
\nabla \uu + \nabla \uu^T$ of the single-layer potential~\eqref{eqn:SLP}
are
\begin{align*}
  \uu(\xx) &= \SS_\uu[\ff](\xx) = \frac{1}{4\pi\mu}\int_{\gamma} \left( 
    -\log \rho + \frac{\rr \otimes \rr}{\rho^2} \right) 
    \ff ds_\yy, \\
    p(\xx) &= \SS_p[\ff](\xx) = \frac{1}{2\pi} \int_{\gamma} 
    \frac{\rr \cdot \ff}{\rho^2} ds_\yy, \\
    T(\xx) &= \SS_T[\ff](\xx) = -\frac{1}{\pi}\int_{\gamma}
      \frac{\rr \cdot \ff}{\rho^2} 
      \frac{\rr \otimes \rr}{\rho^2} ds_\yy,
\end{align*}
where $\rr = \xx - \yy$, and $\rho = \|\xx - \yy\|$.  To compute the
stress tensor on a vesicle, there is a jump in the layer potential
$\SS_T$, and the stress tensor is
\begin{align*}
  T(\xx) = \frac{1}{2}(\nn \otimes \ff) + \frac{1}{2} \left(
    \tt \otimes \left[
    \begin{array}{cc}
      2 t_x t_y & t_y^2 - t_x^2 \\
      t_y^2 - t_x^2 & -2 t_x t_y
    \end{array}
    \right] \ff \right) + 
  \SS_T[\ff](\xx), \quad \xx \in \gamma,
\end{align*}
where $\nn$ and $\tt$ are the vesicle's outward unit normal and tangent
vector, respectively.  With the stress tensor computed for $\xx \in
\gamma$, the shear and normal stresses are
\begin{align*}
  \tau^\tt = (T \nn) \cdot \tt, \qquad
  \tau^\nn = (T \nn) \cdot \nn.
\end{align*}

>>>>

The Adhesion potential for a vesicle membrane has been modeled by a
Lennard-Jones (L.-J.)~type potential
%Sukumaran and Seifert~\cite{suk-sei2001} consider adhesion between a
%single three-dimensional vesicle and a solid wall at $z=0$.  They use
%the adhesion potential of the function form
\begin{align}
\label{eq:adhesion_potential}
  \phi(z) = \mathcal{H} \left[ 
    \left(\frac{\delta}{z}\right)^m - \frac{m}{n} \left(\frac{\delta}{z}\right)^n \right],
\end{align}
where $\mathcal{H}$ is the Hamaker constant and $\delta$ is the adhesion
length scale, and $z$ is the distance between a patch on the vesicle
membrane and a point on the other object, which could be another vesicle
membrane or a flat solid wall.  The characteristics of the
L.-J.~potential are summarized in the left panel of
Figure~\ref{fig:adhesionModel}: The adhesive force is zero at the
equilibrium distance $z=\delta$.  For large distances ($z> \delta$) the
adhesive potential is attractive while for small distances ($z <
\delta$) the potential is repulsive to prevent physical contact.

The exponents $(m,n)$ depend on the geometry and the molecular details
of the two objects under
adhesion~\cite{Book_IntermolecularSurfaceForces}: $(m,n)=(4,2)$
corresponds to two flat, planar surfaces interacting with each other,
and has been a common choice for studying membrane-solid adhesion
\cite{Seifert1990_PRA,suk-sei2001, ShiFengGao2006_ActaMechSin,
LinFreund2007_IntJSolidsStructures, BlountMiksisDavis2013_PRSa,
YoungStone2017_PRF}.  On the other hand, $(m,n) = (12,6)$ corresponds to
the L.-J.~potential between two molecules, and has been used to model
membrane-membrane adhesion  \cite{FlormannAouane2017_SciReports}.  In
our case, the adhesion potential in the integral of
equation~\eqref{eqn:adhesionForce} is between two small patches of lipid
bilayer membrane (because the lipid molecules are coarse-grained in the
continuum modeling).  Thus a reasonable choice for $(m,n)$ between two
coarse-grained membrane patches would be between those for two planar
surfaces and two point molecules.

Large value of $m$ corresponds to a sharp increase in the repulsion
force as two objects are within the separation distance $\delta$.  This
poses a numerical challenge since the problem becomes stiff (i.e.,
requires a very small time step) for large $m$.  The adaptive
time-stepping BIE scheme makes it possible to simulate vesicle adhesive
dynamics with specified numerical precision for reasonable computation
time.  We explore several combinations of $(m,n)$ in the simulations of
two vesicles forming a doublet in a quiescent flow and in a shear flow.
We found that, as long as the close-range interaction is well resolved
both in space (using the near-singular evaluations) and time (using the
high-order adaptive time integration), there is very little difference
in both the dynamic evolution and equilibrium configuration between
$(m,n)=(8,6)$ and $(m,n)=(4,2)$.  Therefore, in this work, we use $(m,n)
= (4,2)$ to regulate the numerical stiffness introduced by the adhesive
force.

Focusing on intermediate adhesive strengths, we use Hamaker constants
ranging from $\mathcal{O}(10^{-1})$ to $\mathcal{O}(10^0)$ times the
bending rigidity modulus.  We assume that the adhesion force from
equation~\ref{eq:adhesion_potential} applies between all pairs of points
on different vesicles, and in Appendix~\ref{sec:appendixA} the net
adhesive force at a point $\xx$ on vesicle $j$ is shown to be
\begin{align}
  \AA\xx:=-\mathcal{H} m \delta^{n}\sum_{\substack{k=1 \\ k \neq j}}^M 
  \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^{m+2}} 
  \left(\delta^{m-n} - \|\xx - \yy\|^{m-n} \right) ds_\yy.
  \label{eqn:adhesionForce}
\end{align}
This summation of adhesive forces is illustrated in the middle panel of
Figure~\ref{fig:adhesionModel}.  The right panel of
Figure~\ref{fig:adhesionModel} is an example of the calculated adhesive
force projected onto their center-of-mass vector. We notice that the
adhesive force is repulsive in the contact region, while for the rest of
the membranes there is an attractive force that acts to pull the two
vesicles together. The summary of boundary integral formulation can be
found in Appendix~\ref{sec:AppendixB}.

@@

To characterize the rheological properties of the suspension of a
doublet, it is necessary to compute the shear and normal stresses.  At a
point $\xx \in \RR^2 \backslash \gamma$ in the fluid bulk, the velocity
$\uu(\xx)$, pressure $p(\xx)$, and total stress tensor $T = - p I +
\nabla \uu + \nabla \uu^T$ of the single-layer potential~\eqref{eqn:SLP}
are
\begin{align*}
  \uu(\xx) &= \SS_\uu[\ff](\xx) = \frac{1}{4\pi\mu}\int_{\gamma} \left( 
    -\log \rho + \frac{\rr \otimes \rr}{\rho^2} \right) 
    \ff ds_\yy, \\
    p(\xx) &= \SS_p[\ff](\xx) = \frac{1}{2\pi} \int_{\gamma} 
    \frac{\rr \cdot \ff}{\rho^2} ds_\yy, \\
    T(\xx) &= \SS_T[\ff](\xx) = -\frac{1}{\pi}\int_{\gamma}
      \frac{\rr \cdot \ff}{\rho^2} 
      \frac{\rr \otimes \rr}{\rho^2} ds_\yy,
\end{align*}
where $\rr = \xx - \yy$, and $\rho = \|\xx - \yy\|$.  To compute the
stress tensor on a vesicle, there is a jump in the layer potential
$\SS_T$, and the stress tensor is
\begin{align*}
  T(\xx) = \frac{1}{2}(\nn \otimes \ff) + \frac{1}{2} \left(
    \tt \otimes \left[
    \begin{array}{cc}
      2 t_x t_y & t_y^2 - t_x^2 \\
      t_y^2 - t_x^2 & -2 t_x t_y
    \end{array}
    \right] \ff \right) + 
  \SS_T[\ff](\xx), \quad \xx \in \gamma,
\end{align*}
where $\nn$ and $\tt$ are the vesicle's outward unit normal and tangent
vector, respectively.  With the stress tensor computed for $\xx \in
\gamma$, the shear and normal stresses are
\begin{align*}
  \tau^\tt = (T \nn) \cdot \tt, \qquad
  \tau^\nn = (T \nn) \cdot \nn.
\end{align*}
